### 2018-01-23 submission/manuscript.Rmd
fix abstract and other grammar errors

**Conclusions.** This data provides support for the importance of the bacterial community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma. One reason for this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

@@

Colorectal cancer (CRC) is a growing world-wide health problem [@siegel_cancer_2016] in which the microbiota has been purported to play an active role in disease pathogenesis [@flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying biomarkers or individual microbes but did not critically investigate how the community changes in CRC.

Although there has been an intense focus on microbiota based biomarker discovery for CRC, the candidate genera seem to be numerous. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, *Escherichia/Shigella* and *Streptococcus* also have good *in vivo* evidence that individual species within the genus can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

@@

Targeting the identification of biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free. Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to the community on the adenoma or carcinoma. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the tissue directly may provide clearer answers but is not without limitations. Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014]. Due to these many differences that could arise between stool and tissue, one question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.

@@

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor or if select members can recapitulate the same information for both adenoma and carcinoma. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the top 5 increased and top 5 decreased RR genera combined resulted in better model classification, based on the area under the curve (AUC). Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification. Finally, we also examined at what effect and sample size the studies that were used were powered for and the sample size needed to get to an 80% power. Our power analysis suggests that each individual study used was underpowered for detecting effect size differences of 10% or below between the case and control groups.

@@

***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
Power transforming and Z-score normalizing OTU richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess differences in broad scale community metrics as disease severity worsens. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether these detectable differences in community resulted in significant increases in RR of lesion (adenoma or carcinoma). For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased RR of Lesion:***
Based on the small increase in RR using $\alpha$-diversity metrics, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue. To investigate this we analyzed all common genera across each study, in tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. For both stool and tissue none of the genera in the top 5 most significantly increased RR were the same between adenoma and carcinoma [Figure 3]. Mouth-associated genera were in the top 5 genera associated with an increased RR of adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and tissue. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and tissue [Figure 3]. Overall, there was little direct overlap of the top 5 increased or decreased RR genera between stool and tissue.

@@

***Whole Community Models Add Important Context:***
Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera based on RR and P-value significance and tested this using two approaches. The first approach used genus level data and tested whether there were any differences in AUC between all genera and selected genera during training on one study or testing on all other studies. The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model AUC across studies when comparing all OTUs versus only OTUs that taxonomically classified to selected genera.

@@

In stool the most common genera in the full genera-based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. Although all four CRC-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B]. For the full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion. Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma. These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for the community in which these microbes exist can increase the ability of models to make predictions.

The data presented herein supports the importance of select genera for carcinoma but not necessarily adenoma formation. Our observations show that when only using the select genera to create models for carcinoma the AUC are similar to using all genera or OTUs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes, within the GI community, are crucial for carcinoma formation. Importantly, it suggests that there may be key members of the community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may need to be used to identify members of the community that are associated with adenoma. Regardless of sample type and whether a full or select model was used our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion. Generally, the full OTU-based models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the full CRC models [Figure S7B-C & E-F]. The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models, is worrying because it is commonly a marker of contamination. The low bacterial biomass of tissue samples coupled with potential contamination could ultimately explain why the these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though specific genera associated with species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data, across study, to the genus level, it is not possible to clearly delineate what specific species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, for stool, the full RF models consistently identified the genus *Bacteroides* as well as OTUs that classified as *Bacteroides* to be important model component across studies. Suggesting that even though *Bacteroides* may not increase the RR of lesion and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionaly, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated micobe and our study clearly shows that regardless of whether the sample is tissue or stool, mouth-associated genera are commonly associated with an increased RR of both adenoma and carcinoma.

The associations between the microbiota and adenoma are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. A small effect size may well be the scope in which differences consistently occur in adenoma based on the results observed within our meta-analysis. Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to poly formation. In contrast to adenoma, our observations suggest that most studies analyzed have sufficient power to detect many changes in carcinoma because of large effect size differences between cases and controls [Figure 7].

>>>>

**Conclusions.** This data provides support for the importance of the bacterial community to both adenoma and carcinoma genesis. The evidence collected within this study on the role of the microbiota in CRC pathogenesis shows stronger associations between carcinoma then adenoma and this may be in part due to the low power to detect more subtle changes in the majority of studies that have been performed to date.

@@

Colorectal cancer (CRC) is a growing world-wide health problem in which the microbiota has been purported to play an active role in disease pathogenesis [@siegel_cancer_2016; @flynn_metabolic_2016]. Numerous studies have shown the importance of both individual microbes [@goodwin_polyamine_2011; @abed_fap2_2016; @arthur_intestinal_2012; @kostic_fusobacterium_2013; @wu_human_2009] and the overall community [@zackular_manipulation_2016; @zackular_gut_2013; @baxter_structure_2014] in tumorgenesis using mouse models of CRC. There have also been numerous case/control studies investigating the microbiota in the formation of both adenoma and carcinoma. A recent meta-analysis investigated whether specific biomarkers could be consistently identified using multiple data sets [@shah_leveraging_2017]. This meta-analysis focused on identifying biomarkers or individual microbes but did so on a small total number of individuals and only investigated stool.

Although there has been an intense focus on microbiota-based biomarker discovery for CRC, the candidate genera seem to be numerous. Some studies point towards mouth-associated genera such as *Fusobacterium*, *Peptostreptococcus*, *Parvimonas*, and *Porphyromonas* as key enriched genera [@ahn_human_2013; @baxter_microbiota-based_2016; @chen_human_2012; @wang_structural_2012; @kostic_fusobacterium_2013; @burns_virulence_2015; @zeller_potential_2014; @flemer_tumour-associated_2017]. Yet, even in some of these studies mouth-associated genera are far from the only microbes identified to be associated with CRC. These other genera include, but are not necessarily limited to, *Providencia*, *Mogibacterium*, *Enterococcus*, *Escherichia/Shigella*, *Klebsiella*, and *Streptococcus* [@burns_virulence_2015; @chen_human_2012; @wang_structural_2012]. In fact, *Escherichia/Shigella* and *Streptococcus* also have good *in vivo* evidence that individual species within the genus can be important in the pathogenesis of CRC [@ecoli_Arthur_2014; @arthur_intestinal_2012; @strep_Aymeric_2017]. Other studies have also identified *Akkermansia muciniphila* and *Bacteroides fragilis* as potential additional markers of CRC with *Bacteroides fragilis* having good mechanistic studies into how this may occur [@weir_stool_2013; @wang_structural_2012; @bfrag_Boleij_2014]. Still, other studies have identified genera like *Roseburia* as increased in CRC while others have reported it to be decreased or have found no difference in any genera at all [@geng_diversified_2013; @wang_structural_2012; @flemer_tumour-associated_2017; @dejea_microbiota_2014].

@@

Targeting the identification of biomarkers within stool seems logical since it offers an easy and cost-effective way to stratify risk and the current gold standard for diagnosis, a colonoscopy, can be time-consuming and is not risk free. Although stool represents an easy and less invasive way to assess risk, it is not clear how reflective this sample is to the community on the adenoma or carcinoma. Some studies have tried to assess this in health and disease but are limited by their sample size [@flemer_tumour-associated_2017; @Flynn_preprint_2017]. Sampling the tissue directly may provide clearer answers but is not without limitations. Due to bowel prep the communities left for sampling may not be reflective of the resident microbiota, but rather a collection of what is able to keep adhered to the mucosa. Additionally, these samples contain more host DNA, potentially limiting the types of analysis that can be done. It is well known that low biomass samples can be very difficult to work with and results can end up being study dependent due to the randomness of contamination [@Salter_contamination_2014]. Due to these many differences that could arise between stool and tissue, an additional question our meta-analysis aims to answer is whether there are consistent patterns that emerge across studies regardless of whether they used stool or tissue samples.

@@

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor or if select members can recapitulate the same information for both adenoma and carcinoma. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the top 5 increased and top 5 decreased RR genera combined resulted in better model classification, based on the area under the curve (AUC). Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification. Finally, we also examined at what effect and sample size the studies that were used were powered for and the sample size needed to get to an 80% power.

@@

***Lower Bacterial Diversity is Associated with Increased RR of Carcinomas:***
Power transforming and Z-score normalizing Operational Taxonomic Unit (OTU) richness, evenness, and Shannon diversity for the entire data set allowed for the combination of data from all studies to assess differences in broad scale community metrics as disease severity worsens. Using linear mixed-effect models to control for study and variable region we assessed whether OTU richness, evenness, or Shannon diversity changed in a step-wise manner with disease severity. In stool, there was a significant decrease in both evenness and Shannon diversity from control to adenoma to carcinoma (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))` and `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`, respectively) [Figure 1A]. We next tested whether these detectable differences in community resulted in significant increases in RR of lesion (adenoma or carcinoma). For fecal samples, a decrease versus the overal median in evenness resulted in a significantly increased RR for carcinoma  (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`),  P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) and adenoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) while a decrease versus the overal median in Shannon diversity only increased the RR for carcinoma (RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` (`r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_lb"], digits = 2))` - `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "ci_ub"], digits = 2))`), P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2]. Using the Bray-Curtis distance metric and PERMANOVA, it was also possible to identify significant bacterial community changes, in specific studies, for both carcinoma versus control and adenoma versus control [Table S1 & S2].

@@

***Mouth-Associated Genera are Associated with an Increased RR of Lesion:***
Based on the small increase in RR using OTU richness and Shannon diversity, we assessed if being higher than the median for any specific genera resulted in an altered RR for adenoma or carcinoma in stool and tissue. To investigate this we analyzed all common genera across each study, in tissue or stool, and assessed whether a relative abundance higher than the median results in an increase or decrease in RR. For both stool and tissue none of the genera in the top 5 most significantly increased RR were the same between adenoma and carcinoma [Figure 3]. Mouth-associated genera were in the top 5 genera associated with an increased RR of adenoma (*Pyramidobacter* [Figure 3A] and *Rothia* [Figure 3C]) and carcinoma (*Fusobacterium*, *Parvimonas*, *Porphyromonas*, and *Peptostreptococcus* [Figure 3B] and *Fusobacterium* [Figure 3D]) for both stool and tissue. Conversely, genera commonly associated with a normal gastrointestinal tract were correlated with a decreased RR for both adenoma and carcinoma for both stool and tissue [Figure 3]. Overall, there was little direct overlap of the top 5 increased or decreased RR genera between stool and tissue.

@@

***Select Genera Models can Recapitulate Whole Community Models:***
Since specific genera increased RR for carcinoma over diversity metrics we assessed whether the bacterial community was better at classifying disease versus only a select group of genera based on RR and P-value significance and tested this using two approaches. The first approach used genus level data and tested whether there were any differences in AUC between all genera and selected genera during training on one study or testing on all other studies. The second approach used OTU level data and tested whether there was a generalized decrease in the 10-fold cross validation (CV) model AUC across studies when comparing all OTUs versus only OTUs that taxonomically classified to selected genera.

@@

In stool, the most common genera in the full genera-based models belonged to genera such as *Ruminococcus*, *Bacteroides*, and *Roseburia* [Figure 6A & B]. Although all mouth-associated genera were present in carcinoma, none were present in the majority of studies and *Fusobacterium* was the only genus present in adenoma [Figure 6A & B]. For the full OTU-based models, *Ruminococcaceae* was present in the top 10 consistently for both adenoma and carcinoma models while *Roseburia* was only present in many adenoma models and *Bacteroides* was present in the overwhelming majority of the carcinoma models [Figure 6C & 6D].

@@

Our study identifies clear differences in diversity, both at the community level and for individual genera, that are present in patients with and without CRC [Figure 1-3]. Although there was a step-wise decrease in diversity from control to adenoma to carcinoma, this did not translate into large effect sizes for the RR of lesion. Even though mouth-associated genera increased the RR of carcinoma, they did not consistently increase the RR of adenoma. These mouth-associated genera are clearly important to carcinoma classification but our observations suggest that accounting for other community members can increase the ability of models to make predictions.

The data presented herein supports the importance of select genera for carcinoma but not necessarily adenoma formation. Our observations show that when only using the select genera to create models for carcinoma the AUC are similar to using all genera or OTUs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes, within the GI community, are crucial for carcinoma formation. Importantly, it suggests that there may be key members of the GI community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may need to be used to identify members of the community that are associated with adenoma. Regardless of sample type and whether a full or select model was used our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

@@

The tissue studies did not provide a clearer understanding of how the microbiota may be associated with lesion. Generally, the full OTU-based models of unmatched and matched [Figure S7E & F] tissue samples were concordant with stool showing that resident microbes were the most prevalent in the top 10 across study. Unlike in stool, *Fusobacterium* was the only mouth-associated bacteria consistently present in the top 10 of the full CRC models [Figure S7B-C & E-F]. The majority of the tissue results seem to be study specific with many top 10 taxa being present only in a single study. Additionally, the presence of genera associated with contamination, within the top 10 most important variables for the genera and OTU models is worrying. The low bacterial biomass of tissue samples coupled with potential contamination could ultimately explain why the these results seem to be more sporadic than the stool results.

One important caveat to this study is that even though specific genera associated with species such as *Bacteroides fragilis* and *Streptococcus gallolyticus* subsp. *gallolyticus* were not identified, it does not necessarily mean these specific species are not important in human CRC [@bfrag_Boleij_2014; @strep_Aymeric_2017]. Since we are limited in our aggregation of the data, across study, to the genus level, it is not possible to clearly delineate what specific species are contributing to overall disease progression. Our observations are not inconsistent with the previous literature on either *Bacteroides fragilis* or *Streptococcus gallolyticus* subsp. *gallolyticus*. As an example, for stool, the full RF models consistently identified the genus *Bacteroides* as well as OTUs that classified as *Bacteroides* to be important model components across studies. Suggesting that even though *Bacteroides* may not increase the RR of lesion and may not vary in relative abundance, like *Fusobacterium*, it is still important in CRC. Additionaly, *Streptococcus gallolyticus* subsp. *gallolyticus* is a mouth-associated micobe and our study clearly shows that regardless of whether the sample is tissue or stool, mouth-associated genera are commonly associated with an increased RR of both adenoma and carcinoma.

The associations between the microbiota and adenoma are inconclusive, in part, because many studies may not be powered effectively to observe small effect sizes. None of the studies analyzed were properly powered to detect a 10% or lower change between cases and controls. A small effect size may well be the scope in which differences consistently occur in adenoma based on the results observed within our meta-analysis. Future studies investigating adenoma and the microbiota need to take power into consideration if we are to reproducibly study whether the microbiota contributes to polyp formation. In contrast to adenoma, our observations suggest that most studies analyzed have sufficient power to detect many changes in carcinoma because of large effect size differences between cases and controls [Figure 7].
